---
title: 'Practical Machine Learning: Course Project'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
by José Carrasquero
Johns Hopkins University Data Science Specialization

##Introduction
In this Course project we will use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants performing one arm dumbell curls in five different ways (correct form, and four different variations with incorrect form):

* A: Correct form (Exactly according to the specification)
* B: Throwing the elbows to the front
* C: Lifting the dumbbell only halfway
* D: Lowering the dumbbell only halfway
* E: Throwing the hips to the front

and train machine learning algorithms able to predict which variation of the exerise is being performed.

##R Packages used:
We will use the following packages:

```{r, warning=FALSE, message=FALSE}
library(caret)
library(rpart)
library(randomForest)
```

## The Data we use:
We proceed to download the training and testing data sets:
```{r cache = TRUE}
if(!file.exists("pml-training.csv")){
        download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", 
                      "/Users/skom/Desktop/coursera/MACHINE_LEARNING_WRITEUP/pml-training.csv",
                      method="curl")
}

if(!file.exists("pml-testing.csv")){
        download.file("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", 
                      "/Users/skom/Desktop/coursera/MACHINE_LEARNING_WRITEUP/pml-testing.csv",
                      method="curl")
}
```

And we proceed to read the csv files into r, in this case we will need to specify that there are 3 different types of NA strings in the files:
```{r cache = TRUE}
df  <- read.csv("pml-training.csv", na.strings = c(NA,"","#DIV/0!"))
df2 <- read.csv("pml-testing.csv" , na.strings = c(NA,"","#DIV/0!"))
```

Then I proceed to eliminate the first seven columns from the training dataset as they contain irrelevant information for the prediction exercise and all columns that have more than 90% of NA values (than we eliminate those same columns from the testing set). Next, the training data set is split into a smaller training and testing data set (60/40 split):
```{r cache = TRUE}
df[1:7]<- list(NULL)
df <- df[,colSums(is.na(df))<nrow(df)*0.9]
set.seed(333)
inTrain <- createDataPartition(y=df$classe,
                               p=0.6, list=FALSE)
training <- df[inTrain,]
testing <- df[-inTrain,]

df2 <- read.csv("pml-testing.csv", na.strings = c(NA,"","#DIV/0!"))
df2<-df2[,names(df2) %in% names(df)]
df2$classe <- 1
```

## Regression tree model
Using the rpart function (from the package of the same name), we proceed to create a regression tree classification model. We see how the prediction model works out against the smaller training and testing datasets (I will only show results for testing sets):
```{r cache = TRUE}
set.seed(333)
model1 <- rpart(classe ~. , data=training)

pred_test1 <-predict(model1,testing, type="class")
testing$predRight <- pred_test1==testing$classe
#table(pred_test1,testing$classe)
confusionMatrix(pred_test1,testing$classe)

pred1 <- predict(model1,df2, type="class")
pred1
```

The regression tree model has an overall accuracy of 73.88% on the testing set (which was subset from the training dataset). The out of sample error is around 26.12%. Looking at the different sensitivities and specificities it is able to predict certain forms of performing the exercise  better than others.

## Random Forest Model
Than using the randomForest function (from the package of the same name), we proceed to create and random forest classification model:
```{r cache = TRUE}
set.seed(333)
model2 <- randomForest(classe ~. , data=training)
model2

pred_test2 <-predict(model2,testing); testing$predRight <- pred_test2==testing$classe
#table(pred_test2,testing$classe)
confusionMatrix(pred_test2,testing$classe)


pred2 <- predict(model2,df2)
pred2
```

This model has a 99.2% accuracy on the training subset and 99.39% accuracy on the testing subset.The out of sample error is around 0.61%.

## Conclusion
For this dataset the Random Forrest model predicts much better in which form the dumbell curl exercise is being performed.
